---
title: "Lipotype pNT Lipid Results"
author: "Robert M Flight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  docx:
    keep-md: true
bibliography: li_ntko.json
csl: plos-computational-biology.csl
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 10,
                      fig.height = 8,
                      dev = c('png', 'svg'),
                      dpi = 600)
```

```{r}
#| label: load-targets
#| include: false
source(here::here("packages.R"))
tar_load(c(eval_lipotype_file, 
           lt_all_annotations,
           lt_sig_cor,
           padjust_limit,
           lt_correlation_sig,
           lt_sig_heatmaps,
           lt_updown_plots))

# # If your chunk output is shown in-line, then you'll need to wrap tar_load()
# # like so:
# 
# withr::with_dir(here::here(), {
#   tar_load(c(target_1, target_2, target_3))
# })
# 
# # This is not needed when using tar_make() to render the document.
```

## Purpose

Two purposes to this report.
1 - Evaluate the reports and results that Lipotype sent (see [here](#Lipotype-Results-Evaluation), and a [summary](#Evaluation-Summary)); 2 - Analyze the Lipotype data ourselves and find some conclusions (see [here](#pNT-to-Lipid-Correlation), and a [summary](#pNT-Correlation-Summary)).

## Lipotype Results Evaluation

```{r}
#| label: include_evaluation
#| child: !expr eval_lipotype_file
```

## pNT to Lipid Correlation

### Methods

#### Extract Fatty-Acids and Replace TAG

From the supplied species level and fatty-acid level data, we created a new data matrix that includes the various individual TAG fatty-acid species.
For example, in the original species level data there is a single entry for TAG 40:0, however; at the fatty-acid level that single TAG 40:0 includes entries with FA of 10:0, 12:0, 14:0, 16:0, and 18:0.
Therefore, in this example, the **single TAG 40:0** will be replaced with **five different FA entries**.

#### Correlation

Information-content-informed Kendall-tau (ICI-Kt) was used to calculate correlation tau and p-values for each combination of pNT and lipid species [@flightInformationContentInformedKendalltauCorrelation2022a].
ICI-Kt allows the inclusion of missing values, assuming that the missing values are the result of truncated normal due to analyte levels being below the level of detection.
Given we expect that this is the most likely reason for values to be missing, it is a reasonable assumption, and therefore using ICI-Kt is appropriate.
Another advantage of ICI-Kt is that it is distribution agnostic, and does not assume that the two variables are related in a linear fashion, merely that the relative pairwise differences between points are similar or dissimilar.
P-values were adjusted for false-discoveries using Bejamini-Hochberg method in {stats::p.adjust} (v 4.2.1) [@benjaminiControllingFalseDiscovery1995; @rcoreteamLanguageEnvironmentStatistical2021a].

#### Annotations & Binomial Enrichment

The lipid annotations we extract include the total carbon chain length and number of double bonds across all fatty-acids, the first and second (if present) fatty-acid chain length and number of double bonds, as well as the combination of lipid class and these values.

For each lipid annotation, we extract the list of lipids that had an adjusted ICI-Kt p-value <= 0.05, and take their correlation values.
We use a binomial test on the number of positive and negative correlated lipid features to evaluate how far from an expected random distribution of 50% positive to 50% negative correlations, as implemented in categoryCompare2 [@flightCategoryCompare22021].
The p-values from the binomial test are adjusted for false-discoveries using the Benjamini-Hochberg method.

#### Calculations

All methods and calculations were implemented in R v 4.2.1 [@rcoreteamLanguageEnvironmentStatistical2021a], with execution managed by {targets} v 0.12.1 [@landauTargetsPackageDynamic2021a].
Data manipulation relied heavily on {dplyr} v 1.0.10 [@wickhamDplyrGrammarData2022].
Some lipid annotations were provided by processing with {lipidr} v 2.11.1 [@mohamedLipidrDataMining2022].
Graphs were generated using {ggplot2} v 3.3.6 [@wickhamGgplot2ElegantGraphics2016].

### Results

#### pNT - Lipid Correlation

Calculating the information-content-informed Kendall-tau (ICI-Kt) correlation of all lipid species to the pNT values results in a distribution that is skewed to the right.
@fig-ici-kt-cor shows the distribution of correlation values observed across all (grey) and for significant (adjusted-p-value <= 0.05) lipids.

```{r}
#| label: fig-ici-kt-cor
#| fig-cap: !expr paste0("All (grey) and significant (black, adjusted-p-value <= ", padjust_limit, ") ICI-Kt correlations of lipid species with pNT across the patients. Red line is the median of all ICI-Kt values.")
ici_kt_lipids = lt_sig_cor |>
  ggplot(aes(x = correlation)) +
  geom_histogram(bins = 100, fill = "lightgrey") +
  geom_vline(aes(xintercept = median(correlation)), color = "red") +
  geom_histogram(data = lt_sig_cor |>
                   dplyr::filter(significant),
                 bins = 100, fill = "black") +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  labs(x = "Lipid - pNT ICI-Kt")
ici_kt_lipids
```

#### Lipid Class Binomial Enrichments

Each lipid was annotated according to various criteria, including lipid class, total and fatty-acid chain specific numbers of carbons and double bonds, and combinations thereof.
For each annotation, we tested the number of positive- and negative-ICI-Kt lipids compared to an expectation of 50/50 positive / negative using a binomial test.
Tables 2 (positive) and 3 (negative) lists the significant annotations (adjusted-p-value <= `r padjust_limit`) by their overall correlation.
The lipid annotations with the most significant adjusted-p-values are those with **18 carbon chain lengths**.
Interestingly, almost **all** significant correlating lipids of several lipid classes all show positive correlations to pNT, including LPC, PC, PC O-, PE, PE O-, and PI.
TAG, while having a mix of positive and negative correlating lipids, is still overwhelmingly positive, with 2X as many positive correlating lipids compared to negative (82 vs 39).
Those lipids that are overwhelmingly negative are the TAG fatty-acids of length 16 and 22, as shown in Table 3.

Table 2. Significantly enriched lipid annotations, overall positive correlations.

```{r}
pretty_positive = lt_correlation_sig |>
  dplyr::filter(overall_cor %in% "positive") |>
  dplyr::arrange(padjust) |>
  format_p_values() %>%
  dplyr::select(annotation, overall_cor, num_positive, num_negative, p, padjust) %>%
  flextable::flextable() %>%
  flextable::colformat_double(digits = 2) %>%
  flextable::set_header_labels(
    overall_cor = "Overall Cor",
    annotation = "Lipid Annotation",
    num_positive = "N Positive",
    num_negative = "N Negative",
    p = "P-Value",
    padjust = "Adjusted P-Value"
  ) %>%
  flextable::set_table_properties(layout = "autofit")
pretty_positive
```
<br>

Table 3. Significantly enriched lipid annotations, overall positive correlations.

```{r}
pretty_negative = lt_correlation_sig |>
  dplyr::filter(overall_cor %in% "negative") |>
  dplyr::arrange(padjust) |>
  format_p_values() %>%
  dplyr::select(annotation, overall_cor, num_positive, num_negative, p, padjust) %>%
  flextable::flextable() %>%
  flextable::colformat_double(digits = 2) %>%
  flextable::set_header_labels(
    overall_cor = "Overall Cor",
    annotation = "Lipid Annotation",
    num_positive = "N Positive",
    num_negative = "N Negative",
    p = "P-Value",
    padjust = "Adjusted P-Value"
  ) %>%
  flextable::set_table_properties(layout = "autofit")
pretty_negative
```

#### Median Correlations by Lipid Class

We can summarize these lipid classes by their median correlations for each lipid class, total (@fig-total-med-cor), TAG (@fig-total-tag-med-cor), and fatty-acid (@fig-fatty-acid-med-cor) specific carbon length and double bond count.

```{r}
#| label: fig-total-med-cor
#| fig-cap: "Total median ICI-Kt correlations for significant lipid species."
n_color = 20
cor_cols = circlize::colorRamp2(seq(-0.1, 0.1, length.out = n_color), scico::scico(n_color, palette = "vik"))
Heatmap(lt_sig_heatmaps$total,
        col = cor_cols,
        name = "ICI-Kt",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_side = "left",
        column_names_side = "top")
```

```{r}
#| label: fig-total-tag-med-cor
#| fig-cap: "Total median ICI-Kt correlations for significant TAG lipid species."
n_color = 20
cor_cols = circlize::colorRamp2(seq(-0.1, 0.1, length.out = n_color), scico::scico(n_color, palette = "vik"))
Heatmap(lt_sig_heatmaps$tag,
        col = cor_cols,
        name = "ICI-Kt",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_side = "left",
        column_names_side = "top")
```

```{r}
#| label: fig-fatty-acid-med-cor
#| fig-cap: "Fatty-acid median ICI-Kt correlations for significant lipid species."
Heatmap(lt_sig_heatmaps$fa, col = cor_cols,
        name = "ICI-Kt",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_side = "left",
        column_names_side = "top")
```

#### Number of Positive-Negative-Correlation

Alternatively, we can look at the number of positively and negatively correlating features for each number of carbons and double bonds.

```{r}
#| label: fig-lpc_class_updown
#| fig-cap: "LPC number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$LPC, nrow = 1)
```

```{r}
#| label: fig-pc_class_updown
#| fig-cap: "PC number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$PC, nrow = 1)
```

```{r}
#| label: fig-pco_class_updown
#| fig-cap: "PC O- number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$`PC O-`, nrow = 1)
```

```{r}
#| label: fig-pe_class_updown
#| fig-cap: "PE number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$PE, nrow = 1)
```

```{r}
#| label: fig-peo_class_updown
#| fig-cap: "PE O- number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$`PE O-`, nrow = 1)
```

```{r}
#| label: fig-pi_class_updown
#| fig-cap: "PI number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$PI, nrow = 1)
```

```{r}
#| label: fig-tag_class_updown
#| fig-width: 12
#| fig-height: 7
#| fig-cap: "TAG number of positive and negative correlating entries by chain length and number of double bonds."
patchwork::wrap_plots(lt_updown_plots$TAG, nrow = 1)
```

### pNT Correlation Summary

* The lipid features overall tend to be more positively correlated with pNT levels across patients.
* The lipid annotations with the most extreme p-values include **fatty-acid chains with 18 carbons** (see Table 2).
  * This is also confirmed in the median correlation heatmaps, where 18:0 and 18:1 have the most positive median correlation values.
  * Further demonstrated in the positive- / negative-count plots of Figure 8 - 14, where the peak in positive seems to commonly be at a fatty-acid chain length of 18.
  * C18:0 likely corresponds to stearic acid, and a common C18:1 fatty-acid is oleic acid.
* There are many classes of lipids where either all or most of the significantly correlating features are positively correlated, including:
  * LPC
  * LPC O-
  * PC
  * PC O-
  * PE
  * PE O-
  * PI
  * TAG
* In contrast to the others, where they are almost all positive, TAG does show a mixture of positively and negatively correlated lipid features, with increasing negatively correlated features with increasing numbers of carbons and double bonds.

## References
