---
title: "Nt KO Quality Control"
author: "Robert M Flight"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  word_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
## target knits Rmds in their own session, so load libraries here.
source(here::here("packages.R"))
lapply(list.files("./R", full.names = TRUE), source)
```


## Purpose

To run quality-assurance / quality-control of the Nt knock-out samples and see if any samples should be removed prior to analysis.

## Analysis

### Load Data

We should have 2 groups with 9 samples each, with Nt WT and KO in Ldlr KO.

```{r load_data}
ko_matrix_single = targets::tar_read("ko_matrix_three")
ko_info = targets::tar_read("ko_info")
targets::tar_load("ko_normalized")
targets::tar_load("sample_manifest")
targets::tar_load("ko_info_block")
flextable::flextable(as.data.frame(table(ko_info$group_name))) %>%
  flextable::autofit()
```


### Principal Components Analysis (PCA)

We decompose the log-transformed values using principal component analysis (PCA), which should give us an idea of how the samples are behaving relative to each other.
Ideally, we should see 4 clusters of the samples.

```{r pca_analysis}
ko_pca = prcomp(t(log1p(ko_matrix_single)), center = TRUE, scale. = FALSE)
ko_pca_info = cbind(as.data.frame(ko_pca$x), ko_info)

ko_pca_var = visualizationQualityControl::visqc_score_contributions(ko_pca$x)

pca_plot = ggplot(ko_pca_info, aes(x = PC1, y = PC2, color = group_name)) + 
  geom_point(size = 3) +
  labs(x = get_pc_label(ko_pca_var, "PC1"),
       y = get_pc_label(ko_pca_var, "PC2"))
pca_plot
```

PC1 is responsible for **97%** of the variance, and it is tied to whether **Ldlr** is knock-out or wild-type (concentrate on the legend for the red and purple dots).

We can confirm this by changing to a 1-dimensional plot of the scores on PC1 (just the x-axis from above) and plotting a histogram of the PC values for each sample and coloring by either Ldlr or Nt genotype.

```{r scores_plot}
scores_plot_nt_1 = ggplot(ko_pca_info, aes(x = PC1, fill = nt_status)) +
  geom_histogram(position = "identity") +
  labs(x = "Sample score on PC1",
       subtitle = "Samples colored by Nt status.")
scores_plot_nt_2 = ggplot(ko_pca_info, aes(x = PC2, fill = nt_status)) +
  geom_histogram(position = "identity") +
  labs(x = "Sample score on PC2",
       subtitle = "Samples colored by Nt status.")
scores_plot_nt_1 / scores_plot_nt_2
```

We can see that plotting the scores for the samples in a histogram, and coloring by Ldlr or Nt gene status, the Ldlr is what is separating the samples from each other, regardless of Nt gene status.

This makes sense, I think, because without Ldlr, then I'm going to assume that LDL is not going to get pulled from the blood.

We can actually test the various groupings of samples, from genotypes to groupings we observed in the sample order (see further down).

```{r test_groups}
ko_info_block2 = dplyr::left_join(ko_info, ko_info_block[, c("sample_id", "block")], by = "sample_id")
full_scores_test = visqc_test_pca_scores(ko_pca$x, ko_info_block2[, c("nt_status", "block")])

ko_pca_info2 = dplyr::left_join(ko_pca_info, ko_info_block[, c("sample_id", "block")], by = "sample_id")

full_scores_test %>%
  dplyr::filter(p.value <= 0.05) %>%
  dplyr::select(PC, variable, p.value)

ko_pca_info2 %>%
  ggplot(aes(x = PC1, y = PC2, color = as.factor(block))) +
  geom_point(size = 2) +
  labs(x = get_pc_label(ko_pca_var, "PC1"),
       y = get_pc_label(ko_pca_var, "PC2"))
```

We can see here that the "block" corresponds very well to the PC's.

### ICI-Kt Correlation

For each sample, we calculate the information-content-informed Kendall-tau (ICI-Kt) correlation with all other smaples.
ICI-Kt allows the missing values to contribute information to the correlation coefficient calculated.
The diagonal represents what fraction of the total metabolites observed are present in an individual sample.
The heatmap displayed is reflected above and below the diagonal.
Any row or column is the correlation of that sample with the other samples.
Ideally, we see blocks of high correlation (greenish - yellow) within samples from the same group, and **lower** correlation between different groups.

```{r correlation}
ko_info = ko_info %>%
  dplyr::mutate(group_short2 = 
                  dplyr::case_when(
      group_short %in% "ntKO_ldWT" ~ "ldWT_ntKO",
      group_short %in% "ntKO_ldKO" ~ "ldKO_ntKO",
      group_short %in% "ntWT_ldWT" ~ "ldWT_ntWT",
      group_short %in% "ntWT_ldKO" ~ "ldKO_ntWT"
                  ))
ko_cor = ici_kendalltau(t(ko_matrix_single))

ko_cor_colors = ggplot2::scale_color_discrete()$palette(2)
names(ko_cor_colors) = unique(ko_info$group_short)

cor_vals = ko_cor$cor

cor_vals = cor_vals[ko_info$sample_id, ko_info$sample_id]
ko_info = as.data.frame(ko_info)
rownames(ko_info) = ko_info$sample_id
cor_order = similarity_reorderbyclass(cor_vals, ko_info[, "group_short", drop = FALSE], transform = "sub_1")

ko_info2 = ko_info[cor_order$indices, ]
cor_vals2 = cor_vals[ko_info2$sample_id, ko_info2$sample_id]
color_list = list(group_short = ko_cor_colors)
row_annotation = HeatmapAnnotation(df = as.data.frame(ko_info2[, c("group_short"), drop = FALSE]),
                                  col = color_list, which = "row",
                                  show_annotation_name = FALSE)
col_annotation = HeatmapAnnotation(df = as.data.frame(ko_info2[, c("group_short"), drop = FALSE]),
                                  col = color_list, which = "column",
                                  show_legend = FALSE, show_annotation_name = FALSE)

cor_map = circlize::colorRamp2(seq(0.8, 1, length.out = 20), viridis::viridis(20))

Heatmap(cor_vals2, col = cor_map, name = "ICI-Kt",
        bottom_annotation = col_annotation,
        right_annotation = row_annotation,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        column_title = "ICI-Kendell-tau Correlation Heatmap",
        column_title_gp = gpar(fontsize = 10))
```

By eye, there do not appear to be any outliers.
Lets also check using a statistical test of the median correlations for each sample by group.

```{r test_outliers}
med_cor = median_correlations(cor_vals, ko_info$group_short)
ko_out = determine_outliers(med_cor)
ggplot(ko_out, aes(x = sample_class, y = med_cor, color = outlier, group = sample_class)) + 
  geom_sina(size = 3) +
  labs(x = "Genotype",
       y = "Median Correlation")
```

The statistical test says there are at least two outliers in the **ld-/nt-** group of samples, and based on these median correlations, the dendrogram below, and the correlation heatmap based on differences below, we will remove the two samples from the **ld-/nt-** group.

### Metabolon Provided Order

Metabolon provided the sample measurements in a particular order, where I noticed that the groups were in contiguous **blocks**.
I'm double checking that if close correlations fall into those same blocks, because if so, that implies that the order they were processed in by Metabolon seems to matter.

```{r check_ko_blocks, fig.height = 8, fig.width = 8}
library(ggraph)

ko_tree = tidygraph::as_tbl_graph(cor_order$dendrogram)
ko_tree_block = ko_tree %>%
  activate(nodes) %>%
  dplyr::left_join(., ko_info_block[, c("sample_id", "group_short", "block")], by = c("label" = "sample_id")) %>%
  dplyr::mutate(block = as.character(block))

ko_tree_block %>%
  ggraph("dendrogram", height = height) +
  geom_edge_elbow2(aes(color = node.block), width = 2) +
  geom_node_text(aes(label = group_short), angle = 90, vjust = 0) +
  coord_cartesian(clip = "off")
```

Figure Z.
Dendrogram of sample - sample distances as (1 - ICI-Kt), and colored by which **block** of samples were **possibly** processed together.
These **blocks** were inferred by the location of samples in the returned Metabolon data sheets.

Figure Z implies that some of the sample grouping is **possibly** due to how they were run in the instrument.
This **assumes** that the order shown in the data sheet is the order in which they were run on the instrument.
We have asked for the timestamp of each sample run to confirm if this is indeed the case.

We would recommend sending the samples in a randomized order, or even just **not** by experimental group.

### Do We Need to Normalize Data?

We can also ask if we need to normalize the data before analysis.
I know Metabolon would claim we don't, but I like to check.

```{r check_to_normalize, fig.width = 16, fig.height = 10}
log_matrix = as.data.frame(log(ko_matrix_single))
long_matrix = tidyr::pivot_longer(log_matrix, cols = everything(), names_to = "sample_id", values_to = "intensity")
long_matrix = dplyr::left_join(long_matrix, ko_info2[, c("sample_id", "group_short2")], by = "sample_id")
long_matrix = long_matrix %>%
  dplyr::filter(!is.na(intensity), !is.infinite(intensity))
long_matrix$sample_id = factor(long_matrix$sample_id, levels = ko_info2$sample_id, ordered = TRUE)

median_matrix = long_matrix %>%
  dplyr::group_by(sample_id) %>%
  dplyr::summarise(intensity = median(intensity))
median_matrix$sample_id = factor(median_matrix$sample_id, levels = ko_info2$sample_id, ordered = TRUE)


ggplot(long_matrix, aes(x = sample_id, y = intensity, color = group_short2, group = sample_id)) + 
  geom_sina() +
  geom_point(data = median_matrix, color = "black", size = 2) +
  theme(axis.text.x = element_text(angle = 90))
```

Yes, we should definitely normalize the data prior to analysis.

### Correlation of Differences

We have elsewhere observed inverted log-fold-changes between the Ldlr WT and Ldlr KO samples.
To evaluate if this is something real, we calculated average log-abundance across all samples after removing missing values, and then calculated differences to the mean, and ICI-Kt correlations across samples, including missing values.

```{r load_differences}
tar_load(ko_diffs_mean)

cor_diff = ici_kendalltau(t(ko_diffs_mean))$cor
ko_info_diff = ko_info %>%
  dplyr::filter(sample_id %in% colnames(cor_diff))
cor_diff = cor_diff[ko_info_diff$sample_id, ko_info_diff$sample_id]
rownames(ko_info_diff) = ko_info_diff$sample_id
diff_order = similarity_reorderbyclass(cor_diff, ko_info_diff[, "group_short", drop = FALSE], transform = "sub_1")

diff_info = ko_info_diff[diff_order$indices, ]
cor_diff2 = cor_diff[diff_info$sample_id, diff_info$sample_id]
diff_color_list = list(group_short = ko_cor_colors)
diff_row_annotation = HeatmapAnnotation(df = as.data.frame(diff_info[, c("group_short"), drop = FALSE]),
                                  col = diff_color_list, which = "row",
                                  show_annotation_name = FALSE)
diff_col_annotation = HeatmapAnnotation(df = as.data.frame(diff_info[, c("group_short"), drop = FALSE]),
                                  col = diff_color_list, which = "column",
                                  show_legend = FALSE, show_annotation_name = FALSE)

diff_map = circlize::colorRamp2(seq(-1, 1, length.out = 20), scico::scico(20, palette = "vik"))

Heatmap(cor_diff2, col = diff_map, name = "ICI-Kt",
        bottom_annotation = diff_col_annotation,
        right_annotation = diff_row_annotation,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        column_title = "ICI-Kendell-tau Correlation Heatmap",
        column_title_gp = gpar(fontsize = 10))
```

Figure X.
ICI-Kt sample-sample correlation using differences to log-mean across samples.

Figure X above indicates that indeed, the Ldlr KO and Ldlr WT are anti-correlated with one another.


### Split PCA

#### Raw Data, Outliers Removed

Because we ended up splitting the data by Ldlr genotype to investigate the effect of the Nt genotype, then there is no reason to keep them together for the PCA.
So lets see what happens in the PCA if we split by Ldlr genotype first.

```{r split_pca}
ko_keep = ko_out %>%
    dplyr::filter(!outlier)
ko_info_pca = ko_info %>%
  dplyr::filter(sample_id %in% ko_keep$sample_id) %>%
  dplyr::left_join(., ko_info_block[, c("sample_id", "block")], by = "sample_id")
split_ldlr = split(ko_info_pca, ko_info_pca$ldlr_status)
  
purrr::iwalk(split_ldlr, function(in_info, info_id){
  tmp_normalize = ko_matrix_single[, in_info$sample_id]
  tmp_normalize2 = t(keep_non_zero_percentage(t(tmp_normalize), sample_classes = in_info$nt_status, keep_num = 0.33, zero_value = 0))
  tmp_pca = prcomp(t(log1p(tmp_normalize2)), center = TRUE, scale. = FALSE)
  tmp_sd = visqc_score_contributions(tmp_pca$x)
  tmp_pca_df = cbind(as.data.frame(tmp_pca$x), in_info)
  
  tmp_anova = visqc_test_pca_scores(tmp_pca$x, in_info[, c("nt_status", "block"), drop = FALSE])
  
  genotype_plot = tmp_pca_df %>%
    ggplot(aes(x = PC1, y = PC2, color = nt_status)) +
    geom_point(size = 2) +
    labs(x = tmp_sd$labels[1], y = tmp_sd$labels[2],
         subtitle = paste0("Ldlr ", info_id))
  block_plot = tmp_pca_df %>%
    ggplot(aes(x = PC1, y = PC2, color = as.factor(block))) +
    geom_point(size = 2) +
    labs(x = tmp_sd$labels[1], y = tmp_sd$labels[2])
  print(paste0("Ldlr ", info_id))
  print(tmp_anova %>%
          dplyr::filter(p.value <= 0.05) %>%
          dplyr::select(PC, variable, p.value))
  print(genotype_plot | block_plot)
})
```

#### Normalized Data, Outliers Removed

The above was all done using the raw data.
Theoretically, centering the data should achieve the same thing as normalizing the data, but lets double check if doing PCA on the normalized data changes anything.

```{r split_pca_normalized}
purrr::iwalk(split_ldlr, function(in_info, info_id){
  tmp_normalize = ko_normalized$normalized_data[, in_info$sample_id]
  tmp_pca = prcomp(t(log(tmp_normalize)), center = TRUE, scale. = FALSE)
  tmp_sd = visqc_score_contributions(tmp_pca$x)
  tmp_pca_df = cbind(as.data.frame(tmp_pca$x), in_info)
  
  tmp_anova = visqc_test_pca_scores(tmp_pca$x, in_info[, c("nt_status", "block"), drop = FALSE])
  
  genotype_plot = tmp_pca_df %>%
    ggplot(aes(x = PC1, y = PC2, color = nt_status)) +
    geom_point(size = 2) +
    labs(x = tmp_sd$labels[1], y = tmp_sd$labels[2],
         subtitle = paste0("Ldlr ", info_id))
  block_plot = tmp_pca_df %>%
    ggplot(aes(x = PC1, y = PC2, color = as.factor(block))) +
    geom_point(size = 2) +
    labs(x = tmp_sd$labels[1], y = tmp_sd$labels[2])
  print(paste0("Ldlr ", info_id))
  print(tmp_anova %>%
          dplyr::filter(p.value <= 0.05) %>%
          dplyr::select(PC, variable, p.value))
  print(genotype_plot | block_plot)
})
```

